Chaotic diffusion of tidal streams

### 2015-05-16 ms.tex
nld review and moved some to results 3

Over long times, a generic orbit in this potential fills the surface of its 3-torus --- that is, the three fundamental frequencies of the orbit are incommensurable. If there exists a relation of the form
\begin{equation}
	\boldsymbol{n} \cdot \boldsymbol{\Omega} = n\, \Omega_1 + m\, \Omega_2 + l\, \Omega_3 = 0
\end{equation}
then the orbit is a resonant orbit and the effective dimensionality of the orbit is decreased to 2D. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits. If an additional resonance relation exists, the orbit is closed and becomes a 1D curve; we refer to such orbits as \emph{bi-resonant} orbits. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. \cite{cincotta06} analytically compute the resonant structure in action-space for a St\"ackel potential. In frequency-space, the resonances are planes, but in action-space are generic 2D surfaces.

The orbit structure of non-integrable potentials can be qualitatively understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $\epsilon$ is a small parameter that determines the perturbation strength \citep[a complete description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99} and the stable regions are surrounded by chaotic layers where stochastic motion occurs. Stochastic or chaotic orbits are characterized by apparently random behavior despite being generated by entirely deterministic systems. The stochastic layers form gaps in action space where tori are destroyed. The orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies evolve with time.

When $\epsilon$ is large, there is no general prediction for the resulting behavior. It seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}.

%For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates \cite[e.g.,][]{goldstein80}. In these coordinates, the actions, $J_n$, are (isolating) integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
%\begin{align}
%	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
%	\act_n &= {\rm const.}\\
%	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}\\
%	\theta_n(t) &= \Omega_n t + \theta_n(0)
%\end{align}
%where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit.  The topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely ``label'' the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus. If a global transformation to angle-action coordinates exists, the potential is said to be \emph{integrable}.
%
%Galactic gravitational potentials will almost certainly not be globally integrable. Dark matter halos of galaxies are expected to be triaxial but even if they are close to axisymmetric or spherical, the superposition of components (e.g., disk, bulge, halo) will break global integrability. If a triaxial (three dof) potential is close to integrable such that the Hamiltonian can be written in terms of a small perturbation, $\epsilon$, away from some integrable Hamiltonian, $H_0$,
%\begin{equation}
%	H(\bs{\theta}, \bs{\act}) = H_0(\bs{\act}) + \epsilon H_1(\bs{\theta}, \bs{\act})
%\end{equation}
%then a large number of regular orbits will survive the perturbation (by way of the KAM theorem)\footnote{3-vectors are represented with bold symbols, e.g., $\bs{J}=(J_1,J_2,J_3)$.}. The tori that survive will in general be separated by regions of irregular or chaotic motion, and thus any transformations to angle-action coordinates can only be defined locally. Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.
%
%The types of motion generic to triaxial systems can be split into several sub-categories (summarized in Figure~\ref{fig:orbit-tree}). Regular motion can be classified by the number of resonance relations obeyed \citep[e.g.,][]{lichtenberg83}: \emph{conditionally periodic} orbits obey no resonance relations, \emph{uni-resonant} orbits obey a single relation of the form $\bs{m}\cdot\bs{\Omega}=0$, and \emph{bi-resonant} orbits obey two resonance relations, $\bs{m}\cdot\bs{\Omega}=0$ and $\bs{n}\cdot\bs{\Omega}=0$, where $\bs{m}$ and $\bs{n}$ are integer vectors.
%
%% Figure 2
%%\begin{figure*}[!h]
%%\begin{center}
%%\includegraphics[width=0.6\textwidth]{figures/orbit-tree.pdf}
%%\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
%%\end{center}
%%\end{figure*}
%
%Truly resonant orbits are rare \citep{merritt99}, but dynamically important in systems with several degrees of freedom as they determine the structure of orbit-space. Stable, resonant orbits generate families of orbits that may behave or appear similarly, whereas unstable resonances are typically associated with regions of chaos. Not all orbits can be trivially associated with resonances due to the infinitely complex, hierarchical structure of resonances (though the importance or ``size'' of a resonance falls off inversely with the norm of its integer vector, $|\bs{m}|^{-\alpha}$, with $\alpha > 0$). In autonomous, two dof systems, stochastic regions are generally surrounded and separated by regular motion, thus chaotic orbits are confined. In three or more dof the chaotic regions will connect and overlap. This is the origin of Arnold diffusion \citep{arnold64}, which allows chaotic orbits to radically change shape by traversing the web of unstable resonances and are only strictly bounded to their own energy hypersurface. This phenomenon is perhaps easier to visualize in action-space: in two dof systems, if a given set of tori are destroyed due to potential perturbations, an orbit in this now chaotic region is still bounded between the surrounding tori due to energy conservation. In three dimensions, the extra degree of freedom allows chaotic orbits to ``escape'' this confinement.
%
%For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic orbits can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale. This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}). 
%
%These local measures quantify the divergence of infinitesimally nearby orbits or the rate of frequency diffusion along an orbit. For an ensemble of orbits with finite spreads in energy or frequencies (e.g., tidal debris), each orbit will have a slightly different characteristic timescale, but we still expect mixing time of the ensemble to correlate with local measures of chaos of the ``parent'' orbit of the ensemble (Section~\ref{sec:ensemble}). The evolution of tidal debris will depend on the ensemble mixing rate around the parent orbit: chaotic mixing will cause the density of the debris to fall off more quickly than the typical power-law behavior expected for regular orbits \citep[e.g.,][]{merritt96, helmi99}. 
%

@@

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic Galactic potentials will have a significant component due to the disk and bulge, may have twisting inertia tensors (cite), radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11, NiHao}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.][]{bailin05}. We expect inclusion of any of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

We use the Dormand-Prince 8th-order Runge-Kutta scheme \citep{prince81} to integrate orbits in the above potential. Specifically, we use a \texttt{Python} wrapper over the \texttt{C} implementation by \cite{hairer93}. For all orbits we ensure that energy is conserved to $\Delta E/E_0 \leq 10^{-8}$ by the end of integration, however most orbits conserve energy to $\Delta E/E_0 \approx 10^{-13}$. Unless otherwise specified the integration timesteps are chosen so that there are 512 steps per strongest orbital period component, but the integrator uses adaptive stepping between each main step in order to satisfy a specified tolerance (we set the absolute tolerance to $\approx$100 times machine precision, $\texttt{atol} = 10^{-13}$).

@@

Figure~\ref{fig:pal5} shows the finite-time Lyapunov exponent computed for the Pal 5 orbit --- shaded region shows the ``converged'' part of the estimate that is used to compute the maximum Lyapunov exponent. By inverting the rate, we find $t_\lambda = XX~{\rm Gyr}$, consistent with our finding in \cite{pearson15}. We then compute the frequency diffusion rate (Equation~\ref{eq:fdrate}) using two consecutive integration windows with length $\approx$50 orbital periods. We find $\fdrate = 10^{-XX}$, which corresponds to a frequency diffusion time, $t_\Omega = XX~{\rm Gyr}$. Though this orbit is computed in a different potential than will be used below, this demonstrates the apparent discrepancy between the timescales over which chaos is estimated to be relevant (XX's of Gyrs) and the integration time over which stream-fanning becomes apparent ($<8$ Gyr). Below, we [...] try to understand by looking at grids of orbits...

>>>>

Over long times, a generic, \emph{conditionally-periodic} orbit in this potential fills the surface of its 3-torus --- that is, the three fundamental frequencies of the orbit are incommensurable. If there exists a relation of the form
\begin{equation}
	\boldsymbol{n} \cdot \boldsymbol{\Omega} = n\, \Omega_1 + m\, \Omega_2 + l\, \Omega_3 = 0
\end{equation}
then the orbit is a resonant orbit and the effective dimensionality of the orbit is decreased to 2D. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits. If an additional resonance relation exists, the orbit is closed and becomes a 1D curve; we refer to such orbits as \emph{bi-resonant} orbits. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. In frequency-space, the resonances are planes, but in action-space are generic 2D surfaces. % \cite{cincotta06} analytically compute the resonant structure in action-space for a St\"ackel potential.

The orbit structure of non-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a complete description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99} and the stable regions are surrounded by chaotic layers where stochastic motion occurs. Stochastic or chaotic orbits are characterized by apparently random behavior despite being generated by entirely deterministic systems. The stochastic layers form gaps in action space where tori are destroyed. The orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies evolve with time. For small values of $\epsilon$, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}.

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori \citep{} --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori caused by the onset of chaotic motion. 
%
%Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.

@@

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic galactic potentials will have a significant component due to the disk and bulge, may have twisting inertia tensors (cite), radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11, NiHao}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.][]{bailin05}. We expect inclusion of any of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

We use the Dormand-Prince 8th-order Runge-Kutta scheme \citep{prince81} to integrate orbits in the above potential. Specifically, we use a \texttt{Python} wrapper over the \texttt{C} implementation by \cite{hairer93}. For all orbits we ensure that energy is conserved to $|\Delta E/E_0| \leq 10^{-8}$ by the end of integration, however most orbits conserve energy to a part in $\approx$$10^{-13}$. Unless otherwise specified the integration timesteps are chosen so that there are 512 steps per strongest orbital period component, but the integrator uses adaptive stepping between each main step in order to satisfy a specified tolerance (we set the absolute tolerance to $\approx$100 times machine precision, $\texttt{atol} = 10^{-13}$).

@@

Figure~\ref{fig:pal5} shows the finite-time Lyapunov exponent computed for the Pal 5 orbit --- shaded region shows the ``converged'' part of the estimate that is used to compute the maximum Lyapunov exponent. By inverting the rate, we find $t_\lambda = XX~{\rm Gyr}$, consistent with our finding in \cite{pearson15}. We then compute the frequency diffusion rate (Equation~\ref{eq:fdrate}) using two consecutive integration windows with length $\approx$50 orbital periods. We find $\fdrate = 10^{-XX}$, which corresponds to a $1\%$ frequency diffusion time, $t_\Omega = XX~{\rm Gyr}$ (the spread in frequencies for a $10^4~\msun$ globular cluster is $\approx$1\%). Though this orbit is computed in a different potential than will be used below, this demonstrates the apparent discrepancy between the timescales over which chaos is estimated to be relevant (XX's of Gyrs) and the integration time over which stream-fanning becomes apparent ($<8$ Gyr). Below, we [...] try to understand by looking at grids of orbits...
